The Companion Mass Distribution to the ELM WD Sample

### 2014-10-10 documents/letter/ms.tex
Updated results/conclusions

Toward these goals, we build the mathematical framework in Section 2 that we employ in our model described in Section 3. Section 4 shows the results of our model, while Section 5 gives a discussion. We conclude in Section 6.

@@

\section{Experiments} \label{sec:experiments}

@@

To test the performance of this Gaussian mixture model, we apply it to three separate sets of mock data (described in detail below), each with 100 ``observed'' systems. These data are generated by randomly drawing the companion mass, $M_2$, from a particular distribution, then using a randomly drawn inclination angle and primary mass, $M_1$, to compute the mass function, $m_f$. We draw primary masses from a uniform distribution, $M_1 \sim \mathcal{U}(0.2,0.4)~\Msun$. We apply the same Gaussian mixture model to all test datasets to infer the parameters of the white dwarf mixture component and neutron star fraction.

In each experiment below, we use an ensemble Markov Chain Monte Carlo algorithm \citep{goodman10} to draw samples from the posterior distribution, $p(\mu_{\rm WD}, \sigma_{\rm WD}, f_{\rm NS} \given \bs{m}_f, \bs{M}_1)$. The algorithm is implemented in the \texttt{Python} programming language \citep{foremanmackey13} and uses an ensemble of individual ``walkers'' to naturally adapt to the geometry of the parameter-space being explored. We run the walkers for an initial, burn-in period of 1000 steps starting from randomly drawn initial conditions (sampled from the priors in Table~\ref{tbl:parameters}). We then re-initialized the walkers from their positions at the end of this run, throw out the burn-in samples, and run again for 1000 steps. We compute the autocorrelation time for each parameter to confirm they have converged.

@@

\subsection{Single Gaussian (white dwarf)} \label{sec:exp1}

@@

\subsection{Two Gaussians (white dwarf + neutron star)} \label{sec:exp2}
For our next test, we use the same Gaussian distribution to generate companion masses for the white dwarf component as the previous example (Section~\ref{sec:exp1}), but we now add an additional neutron star component with parameters
\begin{align}
	\mu_{\rm NS, true} &= 1.4~\Msun\\
	\sigma^2_{\rm NS, true} &= (0.1~\Msun)^2\\
	1.4 < & \frac{M_2}{\Msun} < 2.
\end{align}
We set the neutron star fraction to $f_{\rm NS} = 0.1$. The distribution of mock observations of the mass function, $m_f$, are shown by the middle-left panel of Figure~\ref{fig:tests}. We use MCMC (see description above) to ...

@@

\subsection{Uniform (white dwarf) + Gaussian (neutron star)} \label{sec:exp3}
For a final test, we use the same neutron star component as in Section~\ref{sec:exp2}, but now generate companion masses for the white dwarf component by sampling from a uniform distribution over the range $(0.2,1.2)~\Msun$. We keep the neutron star fraction fixed to $f_{\rm NS}=0.1$. The distribution of mock observations of the mass function, $m_f$, are shown by the middle-left panel of Figure~\ref{fig:tests}. We use MCMC (see description above) to ...

\section{Results}

The ELM WD sample is composed of 55 systems with radial velocity variations fit to orbital solutions, providing precise measurements for $P_{\rm orb}$ and $K$. Fits to spectroscopic templates provide high S/N spectra provide precisely determined $M_1$. We use the spectroscopic solutions found in \citet{gianninas14}. Three systems are eclipsing binaries, with known companion masses (J0345$+$1748: $M_2=0.72~\Msun$ \citep{kaplan14}, J0651$+$2844: $M_2=0.50~\Msun$ \citep{brown11b}, and J0751$-$0141: $M_2=0.97~\Msun$ \citep{kilic14}). These are taken into account in our model through equation {\bf XXX}. Furthermore, six systems in the ELM WD sample show no radial velocity variations, with upper limits of $\approx$20-50 km s$^{-1}$. Some of these systems may be in low inclination angle binaries with radial velocities below the detection limit, or have orbital periods near 24 hours, which are difficult to detect \citep{ELMV}. These LMWDs could also have companions at systematically longer orbital periods, with orbital velocities below the detection limit. Because of the uncertainties in their formation, we restrict these from our sample.

The results of our model are shown in Figure \ref{fig:ELM_post}. The first panel shows the distribution of $M_{2,min}$ for the data set, showing a small high mass component, and the majority of sample with $M_{2,min}$ below $0.8~\Msun$. The second panel shows the posterior distribution of $\mu$ and $\sigma$; The maximum a posteriori Gaussian model has a mean of 0.74 \Msun\ and a standard deviation of 0.26 \Msun. The third panel shows a NS fraction strongly peaked toward zero, however there is a non-negligible tail toward higher NS probabilities. This is further indicated by the fourth panel in Figure \ref{fig:ELM_post} which shows a histogram of NS probabilities for individual systems. These probabilities are obtained by taking the average of the posterior distribution of NS probabilities for each system. Although a few systems have high probabilities, the distribution is strongly clustered near 0\%.

\section{Discussion}

There is no reason to expect the true companion mass distribution to ELM WDs follows a Gaussian curve. However, the posterior distribution in $\mu$ and $\sigma$ is strongly peaked, indicating that our naive Gaussian model is a good descriptor of the companion mass distribution. Our results indicate that the best fit Gaussian for the companions to the ELM WD sample is very similar to that of the population of single DA WDs in SDSS with a mean of 0.6 $\Msun$ \citep{kleinman13}. However, our distribution is significantly wider ($\sigma \approx 0.26~\Msun$ compared with $\sigma \approx 0.1~\Msun$), likely due to mass transfer affecting the mass distribution of the unseen primary WDs as well.

Population synthesis predictions suggest there may be two distinct companion populations to LMWD, composed of both CO and He core WDs. The ratio of these two populations depends on the exact stellar evolution prescriptions assumed \citep[see e.g., ][]{toonen12}. In principle, the ratio of He+He to He+CO WD systems could be determined using two Gaussian model, potentially placing an important constraint on population synthesis models. However adding only one more Gaussian component doubles the number of free parameters from three to six. Finding the maximum likelihood in a three dimensional parameter space is computationally expensive, and moving to a six dimensional space may require more complex numerical routines.  It remains to be seen whether constraints can be placed with 55 measured systems using such a model. We plan to explore this in a future work, as well as a comparison with population synthesis models.

While we were writing this work, we became aware of recent high-speed photometric observations by \citet{hermes14}. Although their search was designed to search 20 LMWD systems for eclipses, they placed limits on the maximum inclination angle of these 20 systems based on the non-detection of eclipses. For eight objects they determined the companion mass based on models fitting the observed photometric variability, primarily due on ellipsoidal variation and doppler beaming. Including these observations in our model is non-trivial and remains outside the scope of this work, however we plan to explore this in a future work.

\begin{figure*}[h!]
\begin{center}
\includegraphics[width=0.95\textwidth]{real-data.pdf}
\caption{Replace this text with your caption}
\label{fig:ELM_post}
\end{center}
\end{figure*}

\subsection{NS Companion Probability}

Systems with NS companions cannot be differentiated from those with WD companions through the mass function and LMWD mass alone. Our model attempts to determine a NS fraction for the population as a whole. This is an improvement over NS probabilities for individual companions calculated through Equation \ref{P_NS_approx}. The third panel in Figure \ref{fig:ELM_post} indicates a NS fraction of a few percent. For 55 systems with observed radial velocity variations, this corresponds to a 1-2 systems with NS companions. However, the top left panel in Figure \ref{fig:tests} shows that even a system with $M_{2,min}=1.0~\Msun$ can be a WD from the high mass wing of our Gaussian distribution. It is not guaranteed that any of the LMWDs in the ELM WD sample have NS companions. Nevertheless, our model indicates a NS fraction of {\bf XXX\%}, corresponding to zero to four systems.

The probability of any particular ELM WD having a NS companion will vary depending on the NS fraction of the sample. This is not taken into account in Equation~\ref{eq:P_NS_approx}. When the NS mass is approximated as a delta function, this can be evaluated analytically, $P(NS)$ reduces to calculating $f_{\rm NS} \times h(M_{\rm NS},m_f,M_1)$:

\begin{equation}
P({\rm NS}) = f_{\rm NS} \frac{(M_1+M_{\rm NS})^{4/3}}{3m_f^{1/3}M_{\rm NS}\sqrt{M_{\rm NS}^2 - \left[m_f(M_1+M_{\rm NS})^2 \right]^{2/3}}} \label{eq:P_NS}
\end{equation}

This equation, approximating the NS population as a delta function around $M=1.4$~\Msun\ instead of the full integral around the Gaussian NS distribution defined in Equation~\ref{eq:P_NS_full}, is accurate to $\sim${\bf XXX}\%. We therefore suggest using this equation, combined with the $f_{\rm NS}$ determined here to provide future estimates for the probability of a NS companion to any LMWD system with a measured $m_f$ and $M_1$.

@@

The fourth panel in Figure {\bf XXX} shows  that only two systems have $P({\rm NS})$ above 25\%: J0811$+$0225 ({\bf XXX}\%) and J1741$+$6526 ({\bf XXX}\%). These systems are ideal candidates for searches for NS companions in radio and X-ray. Radio observations of these objects are in progress, however because a NS companion may be radio quiet or beamed away from us, even if this system is a NS, it may not be detectable in radio. All NS give off blackbody radiation peaked in X-ray wavelengths, therefore X-ray follow-up provides a more stringent test on the potential for a NS companion. Previous searches for LMWDs with X-ray counterparts have all been non-detections, however we show here that all those sources have a low probability of hosting a NS companion. If J0811$+$0225 and J1741$+$6526 also lack X-ray counterparts, then stringent constraints can be placed on $f_{\rm NS}$. In principle, our model can also include the non-detection of radio and X-rays for those systems in which searches have already been performed. We leave this to a future work.

>>>>

Toward these goals, we build the mathematical framework in Section 2 that we employ in our model described in Section 3. We give the results and discussion in Section 4, and conclude in Section 5.

@@

\section{Our Model} \label{sec:experiments}

@@

In our model, we use an ensemble Markov Chain Monte Carlo algorithm \citep{goodman10} to draw samples from the posterior distribution, $p(\mu_{\rm WD}, \sigma_{\rm WD}, f_{\rm NS} \given \bs{m}_f, \bs{M}_1)$. The algorithm is implemented in the \texttt{Python} programming language \citep{foremanmackey13} and uses an ensemble of individual ``walkers'' to naturally adapt to the geometry of the parameter-space being explored. We run the walkers for an initial, burn-in period of 1000 steps starting from randomly drawn initial conditions (sampled from the priors in Table~\ref{tbl:parameters}). We then re-initialized the walkers from their positions at the end of this run, throw out the burn-in samples, and run again for 1000 steps. We compute the autocorrelation time for each parameter to confirm they have converged.

\subsection{Test Cases}
To test the performance of this Gaussian mixture model, we apply it to three separate sets of mock data (described in detail below), each with 100 ``observed'' systems. These data are generated by randomly drawing the companion mass, $M_2$, from a particular distribution, then using a randomly drawn inclination angle and primary mass, $M_1$, to compute the mass function, $m_f$. We draw primary masses from a uniform distribution, $M_1 \sim \mathcal{U}(0.2,0.4)~\Msun$. We apply the same Gaussian mixture model to all test datasets to infer the parameters of the white dwarf mixture component and neutron star fraction.

@@

\subsubsection{Test 1: Single Gaussian (white dwarf)} \label{sec:exp1}

@@

\subsubsection{Test 2: Two Gaussians (white dwarf + neutron star)} \label{sec:exp2}
For our next test, we use the same Gaussian distribution to generate companion masses for the white dwarf component as the previous example (Section~\ref{sec:exp1}), but we now add an additional neutron star component with parameters
\begin{align}
	\mu_{\rm NS, true} &= 1.4~\Msun\\
	\sigma^2_{\rm NS, true} &= (0.1~\Msun)^2\\
	1.4 < & \frac{M_2}{\Msun} < 2.
\end{align}
We set the neutron star fraction to $f_{\rm NS} = 0.1$. The distribution of mock observations of the mass function, $m_f$, are shown by the middle-left panel of Figure~\ref{fig:tests}. We use MCMC (see description above) to ...

@@

\subsubsection{Test 3: Uniform (white dwarf) + Gaussian (neutron star)} \label{sec:exp3}
For a final test, we use the same neutron star component as in Section~\ref{sec:exp2}, but now generate companion masses for the white dwarf component by sampling from a uniform distribution over the range $(0.2,1.2)~\Msun$. We keep the neutron star fraction fixed to $f_{\rm NS}=0.1$. The distribution of mock observations of the mass function, $m_f$, are shown by the middle-left panel of Figure~\ref{fig:tests}. We use MCMC (see description above) to ...

\subsection{The ELM Sample}

The ELM WD sample is composed of 55 systems with radial velocity variations fit to orbital solutions, providing precise measurements for $P_{\rm orb}$ and $K$. Fits to spectroscopic templates provide high S/N spectra provide precisely determined $M_1$. We use the spectroscopic solutions found in \citet{gianninas14}. Three systems are eclipsing binaries, with known companion masses (J0345$+$1748: $M_2=0.72~\Msun$ \citep{kaplan14}, J0651$+$2844: $M_2=0.50~\Msun$ \citep{brown11b}, and J0751$-$0141: $M_2=0.97~\Msun$ \citep{kilic14}). For these systems, our marginal likelihood reduces to:

\begin{equation}
p(m_f \given \theta) = (1-f_{\rm NS}) \mathcal{N}(M_2^* \given \mu_{\rm WD}, \sigma^2_{\rm WD})
\end{equation}

where $M_2^*$ is the mass of the WD companion in these three systems.

Furthermore, six systems in the ELM WD sample show no evidence of orbital motion, with radial velocity upper limits of $\approx$20-50 km s$^{-1}$. Some of these systems may be in low inclination angle binaries with radial velocities below the detection limit, or have orbital periods near 24 hours, which are difficult to detect \citep{ELMV}. These LMWDs could also have companions at systematically longer orbital periods, with orbital velocities below the detection limit. Because of the uncertainties in their formation, we remove these from our analysis.

\begin{figure*}[h!]
\begin{center}
\includegraphics[width=0.95\textwidth]{real-data.pdf}
\caption{The posteriors for our model applied to the ELM WDs. The first panel shows the distribution of $M_2$ for the data set. The black line gives the maximum a posteriori value, while the fainter lines show a set of randomly selected realizations from our model. The second panel shows the posterior distribution of $\mu$ and $\sigma$; The maximum a posteriori Gaussian model has a mean of 0.74 \Msun\ and a standard deviation of 0.26 \Msun. The third panel shows a NS fraction strongly peaked toward zero, however there is a non-negligible tail toward higher NS probabilities. This is further indicated by the fourth panel in Figure \ref{fig:ELM_post} which shows the relation between $M_f$ and $P({\rm NS})$ for individual systems. }
\label{fig:ELM_post}
\end{center}
\end{figure*}

\section{Results and Discussion}

The results of our model are shown in Figure \ref{fig:ELM_post}. The maximum a posteriori Gaussian model has a mean of 0.74 \Msun\ and a standard deviation of 0.26 \Msun. While the well-defined peak in the posterior distribution demonstrates convergence of the model the extended distribution toward larger variances could indicate that the true distribution does not match a Gaussian. However it is interesting that the best fit Gaussian for the companions to the ELM WD sample is similar to that of the population of single DA WDs in SDSS with a mean of 0.6 $\Msun$ \citep{kleinman13}. Our distribution is significantly wider ($\sigma \approx 0.26~\Msun$ compared with $\sigma \approx 0.1~\Msun$), likely due to mass transfer affecting the mass distribution of the unseen primary WDs as well.

Our results are independent of astrophysical expectations since they do not include any priors on $\mu$ and $\sigma$, apart from limits for $M_2$ of 0.2 \Msun (based on observations of LMWDs) and 1.4 \Msun (based on the Chandrasekhar mass). In principle, priors could be added based on population synthesis models which suggest there are two dominant companion populations to LMWD, composed of CO and He core WDs \citep{han98}. The ratio and distribution of these two populations depends on the exact stellar evolution prescriptions assumed \citep[see discussion in e.g., ][]{toonen12}.

Our model could be expanded to include a two Gaussian model one for each WD population. The ratio of these populations could be determined, and potentially place an important constraint on population synthesis models. However adding only one more Gaussian component doubles the number of free parameters from three to six. Finding the maximum likelihood in a three dimensional parameter space is computationally expensive, and moving to a six dimensional space may require more complex numerical routines.  It remains to be seen whether constraints can be placed with 55 measured systems using such a model.

The third panel in Figure~\ref{fig:ELM_post} shows a NS fraction strongly peaked toward zero, however there is a non-negligible tail toward higher NS probabilities. Our model does indicate an upper limit to the NS fraction of {\bf XXX\%} at the 95\% confidence level. For 55 systems with observed radial velocity variations, this corresponds to a {\bf 0-4?} systems with NS companions. These results are in agreement with independent constraints from \citet{vLeeuwen07} who set an upper limit on $P({\rm NS})$ of 18$\pm5$\% based on the non-detection of radio emission from eight LMWD targets. With non-detections of radio emission from an expanded sample of 15 LMWD targets in the ELM WD sample, \citet{agueros09b} place a more stringent upper limit on $P({\rm NS})$ of $\substack{+4 \\ -2}\%$.

Since our results are still consistent with a 0\% NS fraction, it is not guaranteed that any of the LMWDs in the ELM WD sample have NS companions. However the fourth panel in Figure \ref{fig:ELM_post} indicates which ELM WDs have the highest probability of hosting NS companions. Here, we show the relation between $m_f$ and $P({\rm NS})$ for individual systems. This panel shows that $P({\rm NS})$ only starts to significantly deviate from zero for $m_f\approx0.6 \Msun$ Although a few systems have high probabilities, the distribution is strongly clustered near 0\%.

The two systems with $P({\rm NS})$ above 10\% are J0811$+$0225 ({\bf XXX}\%) and J1741$+$6526 ({\bf XXX}\%). These systems are ideal candidates for searches for NS companions in radio and X-ray. Radio observations of these objects are in progress, however because a NS companion may be radio quiet or beamed away from us, a NS companion to these LMWDs may not be detectable in radio \citep{vLeeuwen07}. All NS give off blackbody radiation peaked in X-ray wavelengths, therefore X-ray follow-up provides a more stringent test on the potential for a NS companion \citep{agueros09a}. Previous searches for LMWDs with X-ray counterparts have all been non-detections, however we show here that all those sources have a low probability of hosting a NS companion. If J0811$+$0225 and J1741$+$6526 also lack X-ray counterparts, then even more stringent constraints can be placed on $f_{\rm NS}$. In a future work, we plan to combine our model results here with radio and x-ray non-detections, placing more stringent constraints on $P({\rm NS})$.

It is worth noting that the fourth panels in our test cases in Figure~\ref{fig:tests} shows that our model is fairly good at identifying those systems most likely to host NS's. However it also shows that roughly half of the LMWDs with NS companions may appear with a deceptively low $f_{\rm NS}$. Systems with NS companions cannot all be differentiated from those with WD companions through the mass function and LMWD mass alone. Therefore {\it any} LMWD could host a NS companion.

The probability of any particular ELM WD having a NS companion will vary depending on the NS fraction of the sample. This is not taken into account in Equation~\ref{eq:P_NS_approx}. {\bf TEXT ON USING PYTHON SCRIPT FOR DETERMINING P-NS FOR FUTURE DISCOVERED LMWDs.}

Recently, high-speed photometric observations by \citet{hermes14}. Although their search was designed to search 20 LMWD systems for eclipses, they placed limits on the maximum inclination angle of these 20 systems based on the non-detection of eclipses. For eight objects they determined the companion mass based on models fitting the observed photometric variability, primarily due on ellipsoidal variation and doppler beaming. Including these observations in our model is non-trivial and remains outside the scope of this work, however we plan to explore this in a future work.

@@
